---
title: "Montgomery County, PA - 911 Calls and COVID-19 Exploratory Data Analysis"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, include=FALSE}
library(tidyverse)
library(lubridate)
library(data.table) # need this for the "%like%"function
source("./src/funcs.R")
```

```{r, echo=FALSE}
tz_latest <- read.csv("https://storage.googleapis.com/montco-stats/tz.csv")
jhu_latest <- read.csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_US.csv")
jhu_deaths <- read.csv("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_US.csv")
```


```{r, echo=FALSE}
# data cleanup
tz_latest$datetime <- ymd_hms(tz_latest$timeStamp, tz=NULL)
tz_latest$desc <- as.character(tz_latest$desc)
tz_latest$zip <- as.character(tz_latest$zip)

```
**This is a work in progress. Please check in regularly for more updated plots. I try to update this page at least twice a week or more.**

## Historical Trends In 911 Calls From 2015 To Present For Select Complaints

### Fever
Very notable spike in calls at the beginning of 2020 compared to previous years.  Scroll down to see a more focused plot specific for 2020 calls.
*The ongoing drop in calls is concerning because it may imply people are now reluctant to call 911 and go to hospital for fear of getting exposed to COVID-19.*


```{r, echo=FALSE, warning=FALSE, message=FALSE}

plot_by_complaint(by_complaint(tz_latest, "fever", "EMS: FEVER"), "Fever")

```

### Respiratory Emergencies
A very slight increase here which surprised me. I was expecting to see a much higher spike considering the initial understanding of SARS-CoV2 to be a respiratory virus.  

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Calls for Respiratory Emergencies
plot_by_complaint(by_complaint(tz_latest, "respiratory", "EMS: RESPIRATORY EMERGENCY"), "Respiratory Emergencies")

```

### Cardiac Emergencies
No noticeable spike here around COVID-19. Hoever, the ongoing drop in calls is concerning because it may imply people are now reluctant to call 911 and go to hospital for fear of getting exposed to COVID-19.  This trend is now being reported in the news as well (see references section below)


```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Calls for Cardiac Emergencies
plot_by_complaint(by_complaint(tz_latest, "cardiac", "EMS: CARDIAC EMERGENCY"), "Cardiac Emergencies")

```

### Cardiac Arrests
Fairly consistent, although maybe a bit of a peak in April?

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Calls for Cardiac Arrests
plot_by_complaint(by_complaint(tz_latest, "cardiac_arrests", "EMS: CARDIAC ARREST"), "Cardiac Arrests")

```

### Medical Alert Alarms
A noticeable increase in medic alert alarms starting late 2019, may be completely unrelated but would be interesting to triangulate and see if most of these are from nursing home locations?  

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Calls for Medic Alert Alarms
plot_by_complaint(by_complaint(tz_latest, "medic_alert_alarms", "EMS: MEDICAL ALERT ALARM"), "Medical Alert Alarms")

```


### Traffic-related incidents
A total drop off in 911 calls for traffic-related, likely a strong reflection of the quarantine measures, fewer people out and about.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Calls for traffic incidents
plot_by_complaint(by_complaint_like(tz_latest, "traffic", "Traffic"), "Traffic-Related Incidents")

```


## SINCE START OF COVID-19
Narrowing down the focus on the data to just 2020, from around the presumed start of COVID-19 spread in the US.  While we now see the number of calls dropping to early February levels, the ongoing drop in calls is concerning because it may imply people are now reluctant to call 911 and go to hospital for fear of getting exposed to COVID-19.


```{r, echo=FALSE, warning=FALSE, message=FALSE}

# fever
plot_by_complaint_2020(by_complaint_2020(tz_latest, "fever", "EMS: FEVER"), "Fever")

# Calls for Respiratory Emergencies
plot_by_complaint_2020(by_complaint_2020(tz_latest, "respiratory", "EMS: RESPIRATORY EMERGENCY"), "Respiratory Emergencies")

# Calls for Cardiac Emergencies
plot_by_complaint_2020(by_complaint_2020(tz_latest, "cardiac", "EMS: CARDIAC EMERGENCY"), "Cardiac Emergencies")

# Calls for Cardiac Arrests
plot_by_complaint_2020(by_complaint_2020(tz_latest, "cardiac_arrests", "EMS: CARDIAC ARREST"), "Cardiac Arrests")

```

```{r, echo=FALSE, warning=FALSE,message=FALSE}
# Extract Covid-19 data for Montco PA only
  
montco_cases <- jhu_latest %>% filter(Combined_Key == "Montgomery, Pennsylvania, US") %>%
  select(starts_with("X")) 

montco_data_cases <- montco_cases %>% rbind(montco_cases, as.vector(unlist(str_split(colnames(montco_cases), " ")))) %>%
  t() %>%
  as_tibble(.name_repair = NULL) %>%
  mutate(date = str_sub(V3, 2)) %>%
  select(date, number_of_cases = V1) %>%
  mutate(date =  mdy(date), number_of_cases = as.numeric(number_of_cases)) %>%
  filter(wday(date) == 1) # to matchup with dates in the fevers table


ggplot(montco_data_cases, aes(x=as.Date(date), y=number_of_cases)) +
  geom_bar(stat = "identity") + scale_x_date(date_breaks = "2 weeks") + 
  geom_smooth(method = "gam", color = "red", se = FALSE) +
  theme_minimal() +
  labs(title = "COVID-19 Confirmes Cases by Week", subtitle = "Montgomery County, PA (tuesdays") +
  xlab("Week") +
  ylab("Total Number of Cases") +
  theme(axis.text.x=element_text(angle=60, hjust=1))
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# Extract Covid-19 death data for Montco PA only

montco_deaths <- jhu_deaths %>% filter(Combined_Key == "Montgomery, Pennsylvania, US") %>%
  select(starts_with("X"))

montco_data_deaths <- montco_deaths %>% rbind(montco_deaths, as.vector(unlist(str_split(colnames(montco_deaths), " ")))) %>%
  t() %>%
  as_tibble(.name_repair = NULL) %>%
  mutate(date = str_sub(V3, 2)) %>%
  select(date, number_of_deaths = V1) %>%
  mutate(date =  mdy(date), number_of_deaths = as.numeric(number_of_deaths)) %>%
  filter(wday(date) == 1) # to matchup with dates in the fevers table

#montco_data_deaths
ggplot(montco_data_deaths, aes(x=as.Date(date), y=number_of_deaths)) +
  geom_bar(stat = "identity") + scale_x_date(date_breaks = "2 weeks") + 
  geom_smooth(method = "gam", color = "red", se = FALSE) +
  theme_minimal() +
  labs(title = "COVID-19 Deaths by Week", subtitle = "Montgomery County, PA (tuesdays") +
  xlab("Week") +
  ylab("Total Number of Deaths") +
  theme(axis.text.x=element_text(angle=60, hjust=1))
```


### Comparing Calls for Fever with Confirmed Cases 
Exploring the correlation between number of calls for fever and number of confirmed cases we see a lag in the trends which may be indicative of the delay between when COVID-19 may have started affecting people in the area vs when confirmation of diagnosis (not an individual basis but from a general big-picture perspective)
```{r, echo=FALSE, warning=FALSE}
#merge the two
tz_latest_fever_2020 = by_complaint_2020(tz_latest, "fever", "EMS: FEVER")
tz_latest_fever_2020$week <- as.Date(tz_latest_fever_2020$week)
merged <- left_join(tz_latest_fever_2020, montco_data_cases, by = c("week" = "date"))
merged$number_of_cases[is.na(merged$number_of_cases)] <- 0

```

```{r, echo=FALSE, warning=FALSE}
coeff <- 50

# A few constants
callsColor <- "blue"
casesColor <- "magenta"

ggplot(merged, aes(x=week)) +
  
  geom_line( aes(y=number_of_calls), color=callsColor) +
  scale_x_date(date_breaks = "2 weeks") +
  theme_minimal() +
  geom_line( aes(y=number_of_cases / coeff), color=casesColor) +
  
  scale_y_continuous(
    
    # Features of the first axis
    name = "Number of Calls",
    
    # Add a second axis and specify its features
    sec.axis = sec_axis(~.*coeff, name="Total Confirmed Cases")
  ) + 

  theme(
    axis.title.y = element_text(color = callsColor, size=13),
    axis.title.y.right = element_text(color = casesColor, size=13),
    axis.text.x=element_text(angle=60, hjust=1)
  ) +

  labs(title = "Number of 911 Calls for Fever compared with Number of Confirmed Cases", subtitle = "Montgomery County, PA")
```

### DATA SOURCES:
* COVID-19 Case Numbers - <a href="https://coronavirus.jhu.edu/map.html">John Hopkins University</a>
* Montgomery County EMS Data - Compiled and provided by <a href=https://www.kaggle.com/mchirico/montcoalert>  Mike Chirico</a> 

## REFERENCES:
* <a href="https://www.acc.org/latest-in-cardiology/articles/2020/04/22/13/36/the-new-pandemic-threat-people-may-die-because-theyre-not-calling-911-coronavirus-disease-2019-covid-19">The New Pandemic Threat: People May Die Because They’re Not Calling 911</a>
* <a href="https://www.npr.org/sections/health-shots/2020/05/06/850454989/eerie-emptiness-of-ers-worries-doctors-where-are-the-heart-attacks-and-strokes">Eerie Emptiness Of ERs Worries Doctors: Where Are The Heart Attacks And Strokes?</a>


Last Updated: `r Sys.Date()`
```{r, echo=FALSE}
last_updated <- Sys.Date()

```
